Template for Oxford University Press papers

### 2025-01-23 inst/manuscript/paper_oxford/paper_oxford.Rmd
get oxford style working for Fisheries

---
title: 'hatchR: A toolset to predict hatch and emergence phenology in wild fishes'
tags:
  - R
  - fishes
  - developmental phenology
authors:
  - name: Morgan M. Sparks
    orcid: 0000-0003-0787-2218
    affiliation: 1 # (Multiple affiliations must be quoted)
  - name: Eli Felts
    orcid: 0000-0002-5888-9266
    affiliation: 2
  - name: Bryan M. Maitland
    orcid: 0000-0002-4491-5064
    affiliation: 1
affiliations:
 - name: US Forest Service, Rocky Mountain Research Station, Boise, ID, USA
   index: 1
 - name: US Fish and Wildlife Service, Idaho Fish and Wildlife Conservation Office, Orofino, ID, USA
   index: 2
citation_author: Sparks et al.
date: 07 July 2025
year: 2025
output: rticles::joss_article
csl: apa.csl
journal: JOSS
editor_options:
  chunk_output_type: console
bibliography: ms_refs.bib
---

@@

# Abstract

Understanding the timing of key life history events is necessary for managing and conserving populations. Historically, models to predict hatch and emergence timing for fishes were difficult to employ in wild settings because average incubation temperature was needed as the primary parameter in predictive models. However, recent improvements to these techniques reworked models such that they could be applied in wild environments as long as users had data for when adult fish spawned and a record of average daily temperature over the course of development. Despite these improvements, their application remains limited due to few parameterizations for varying species, being largely limited to salmonids. Here we present **hatchR**, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses. **hatchR** allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package. **hatchR** is also distributed in two forms—an open source R package for maximum customization, as well as an HTML graphical-user-interface web application for individuals not familiar with scripting languages. To demonstrate potential uses, we present two case studies as likely applications for this software. **hatchR** promises to open many exciting avenues in research and management of fishes during their early life history.

@@

```{r echo = FALSE, out.height = '50%', fig.align = 'center', fig.cap= "Full workflow for hatchR. Data processes are demonstrated with beige circles and pakcage functions with white rectangles."}
knitr::include_graphics("flowchart.png")
```

@@

| date       | temperature |
|------------|-------------|
| 2000-01-01 | 2.51        |
| ...        | ...         |
| 2000-07-01 | 16.32       |
| ...        | ...         |
| 2000-12-31 | 3.13        |

@@

```{r}
# create date object for a year with 30 min reading intervals
dates <- seq(from = ymd_hms("2000-07-01 00:00:00"),
             to = ymd_hms("2001-06-30 23:59:59"), length.out = 17568)

@@

```{r, fig.cap= "Full output of the plot_check_temp() visual data check function from hatchR. Note that the red and blue lines are customizable by the user to create outlier thresholds for their visual check."}
plot_check_temp(data = year_sim,
                dates = date,
                temperature = temp,
                temp_min = 0, # temp_min and max lines are 
                              # user customizable
                temp_max = 25)
```

@@

```{r}
# summarize
year_sim_summ <- summarize_temp(data = year_sim,
                                dates = date,
                                temperature = temp)

@@

Users can either use Salmonid models from `model_table` included in the package and the Shiny app. As discussed, these models are included because parameterizations already existed in the literature. However, these parameterizations are limited to Pacific Salmon and Bull Trout. To widen the application of the effective value approach, we include a `fit_model()` function and is species agnostic (as long as develop generally follows the power law.

The function `fit_model()` uses data in which average incubation temperature (°C) and days to phenological event (as two vectors) are the inputs and estimates parameter coeficients for *log~e~a* and *b* using `stats::nls()`

@@

Below, we demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package. We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for fishes beyond the scope of the original effective value approach. These parameterizations are for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*), Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005. We demonstrate parameterization for just Smallmouth Bass here for concision, but the full code set for all species is available at <https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper.Rmd>.

@@

```{r}
###  make temp regime
set.seed(123)
# create random temps and corresponding dates
temps_sim <- sort(rnorm(n =30, mean = 16, sd = 1), decreasing = FALSE)
dates_sim <-  seq(from = ymd("2000-07-01"),
             to = ymd("2000-07-31"), length.out = 30)

@@

```{r}
#model fits
smb_mod$r_squared; cat_mod$r_squared; sturgeon_mod$r_squared

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap= "Days to hatch as predicted using custom parameterizations for Channel Catfish (blue), Lake Sturgeon (black, and Smallmouth Bass (green) as generated by the fit_model() function in hathcR. Models are all run using a spawn date starting June 1 for a randomly generated ascending thermal regime with mean temperature 16 °C"}
### predict_phenology

@@

```{r}
sockeye_hatch_mod <- model_select(
  author = "Beacham and Murray 1990", 
  species = "sockeye", 
  model = 2, 
  development_type = "hatch"
  )
```

@@

```{r, warning=FALSE}
WI_hatch <- predict_phenology(
  data = woody_island,
  dates = date,
  temperature = temp_c,
  spawn.date = "1990-08-18",
  model = sockeye_hatch_mod
  )
```

@@

```{r}
WI_hatch$days2done
WI_hatch$dev.period
```

@@

```{r, eval=FALSE}
str(WI_hatch)
```

@@

```{r, fig.cap= "Output of the plot_phenology() function from hatchR. The function plots the output of the predict_phenology() function and displays time to develop, as well as the ambient thermal regime (blue), scaled daily effective value (green), and the scaled cumulative sum of effective values (orange)."}
plot_phenology(WI_hatch)
```

@@

A common management scenario where developmental phenology might be useful would be trying to understand if fish might be free-moving before some management action. For instance, will have fish have emerged from redds when a stream section has been opened to grazing or bridge decommissioning will commence?

In this scenario, we will consider the grazing example and Bull Trout, a threatened fish in the United States under the Endangered Species Act [@nolfi2024], and the East Fork Salmon River, a key Bull Trout population in the upper Salmon River watershed. The fisheries manager there wants to know if fish will likely be out of the gravel and free-swimming by June 1st. In this system, it is expected that Bull Trout will be done spawning by the end of September, so we'll consider the last possible spawn date as September 30th.

We demonstrate this first case study using the graphical user interface portion of the **hatchR** ecosystem found at <https://elifelts.shinyapps.io/hatchR/>. Users will first upload their data with the `Import Data` window, which requires them to select their file on their personal computer, provide the program with the columns corresponding for dates and temperatures, and then provide the format in which dates are coded (*e.g.*, year-month-day or day-month-year). Once data is uploaded the program automatically plots the user's data using `plot_check_temp()` in the background and provides them the outputted graphical check. After uploading and checking data, the user switches to the `Model Phenology` window. In this circumstance, we use the preloaded parameterization for bull trout from @austin2019 with the `Existing` button for model selection, which the user selects with the various drop down options in the menu. After the model is selected, the user can choose multiple spawn dates from the interactive calendar provided. We show results for spawning for September 30th as indicated in the example above. Once dates are chosen, a table entry for each spawn date is outputted in the `Phenology Summaries` tab and corresponding plot with data from each spawn date in the `Timeline Plot` tab. Output from predicting phenology and the resulting figure are downloadable from their respective tabs. The process is demonstrated in full in Figure XXX, but the interface is described more completely in the Articles on **hatchR**'s website <https://bmait101.github.io/hatchR/>.

@@

```{r}
hatch_res <- isaak_summ_bt |> 
  mutate(
    dev_period = map2(
      summ_obj, spawn_dates, # map across our site object and spawn dates
      predict_phenology,
      temperature = daily_temp,
      model = bt_hatch,
      dates = date
      ) |> 
      map_df("dev.period") |> # pull out just dev.period results
      list()
    ) |> 
  select(site, dev_period) |> # just select the columns we want
  unnest(cols = c(dev_period)) |> # unnest everything
  mutate(days_to_hatch = stop - start) # make a new column of days to hatch
```

```{r, message=FALSE, echo=FALSE, fig.cap= "Densities of predicted days to hatch for Bull Trout for 139 putative spawning locations across Idaho for four years of shared temperature data. Early spawning considers if a parent spawned September 1, Peak-September 15, and Late-September 30."}
cut_ints <- date(
  c(
    "2011-08-31", "2012-08-31",
    "2013-08-31", "2014-08-31",
    "2015-08-31"
  )
)

@@

# Acknowledgements

We thank Laura Koller for her help designing the **hatchR** logo. Dan Isaak provided useful discussion about model development and temperature datasets.

>>>>

---
title: Template for Oxford University Press papers
author:
  - name: Morgan M. Sparks
    email: Morgan.Sparks@usda.gov
    orcid: 0000-0003-0787-2218
    corresponding_author: TRUE # Automatic footnote using corresponding_author
    affiliation: A
  - name: Eli Felts
    orcid: 0000-0002-5888-9266
    affiliation: B
  - name: Bryan M. Maitland
    orcid: 0000-0002-4491-5064
    affiliation: A
    footnote:
address:
  - code: A
    department: Rocky Mountain Research Station
    organization: U.S. Forest Service
    #postcode: 
    city: Boise
    state: Idaho # Mix and match elements to get proper address
    country: United States
  - code: B
    department: Idaho Fish and Wildlife Conservation Office
    organization: U.S. Fish and Wildlife Service
    city: Orofino
    state: Idaho # Mix and match elements to get proper address
    country: United States
abstract:
  - "Understanding the timing of key life history events is necessary for managing and conserving populations. Historically, models to predict hatch and emergence timing for fishes were difficult to employ in wild settings because average incubation temperature was needed as the primary parameter in predictive models. However, recent improvements to these techniques reworked models such that they could be applied in wild environments as long as users had data for when adult fish spawned and a record of average daily temperature over the course of development. Despite these improvements, their application remains limited due to few parameterizations for varying species, being largely limited to salmonids. Here we present **hatchR**, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses. **hatchR** allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package. **hatchR** is also distributed in two forms—an open source R package for maximum customization, as well as an HTML graphical-user-interface web application for individuals not familiar with scripting languages. To demonstrate potential uses, we present two case studies as likely applications for this software. **hatchR** promises to open many exciting avenues in research and management of fishes during their early life history.
"
keywords:
  - developmental phenology
  - effective value model
  - R
  - Shiny
#boxedlist: # Not fully compatible with all document formats
#  - First point
#  - Second point
competing_interests: |
  There are no competing interest.
author_contributions: |
  MMS and BMM designed and wrote the package. EF wrote the Shiny app. All authors wrote and edited the manuscript.
acknowledgments: |
  We thank Laura Koller for her help designing the **hatchR** logo. Dan Isaak provided useful discussion about model development and temperature datasets.

bibliography: ms_refs.bib # Includes refs in OUP example template
## When using `citation_package = "default"`, a CSL file can be used
#csl: https://www.zotero.org/styles/oxford-university-press-note
output:
  rticles::oup_article:
    oup_version: 1 # 1 = 2020 CTAN OUP CLS package; 0 = 2009 OUP CLS package
    journal: "Fisheries"
    document_style: "contemporary" # Can be contemporary, modern, traditional
    papersize: "large" # Can be large, medium, small
    citation_package: "default" # Uncomment when using a CSL; default "natbib"
    namedate: TRUE # Set FALSE to use numeric refs; Default FALSE
    #number_sections: FALSE # Uncomment to not number sections; default TRUE
    #number_lines: TRUE # Use `lineno` package - Default FALSE
    #number_lines_options: ["mathlines","switch"]  # Options for latex lineno package.
    #onecolumn: TRUE # Uncomment for one column format; default FALSE
    extra_dependencies:
      - booktabs # to use with knitr::kable() example below

## Example of pandoc's variable
#urlcolor: orange
#linkcolor: green
#citecolor: red
header-includes:
  #- \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
  - \theoremstyle{thmstyleone} # Theorem stuff from OUP template
  - \newtheorem{theorem}{Theorem} #  meant for continuous numbers. %%\newtheorem{theorem}{Theorem}[section] # meant for sectionwise numbers. optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
  - \newtheorem{proposition}[theorem]{Proposition} # %%\newtheorem{proposition}{Proposition}" # to get separate numbers for theorem and proposition etc.
  - \theoremstyle{thmstyletwo}
  - \newtheorem{example}{Example}
  - \newtheorem{remark}{Remark}
  - \theoremstyle{thmstylethree}
  - \newtheorem{definition}{Definition}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, # By default, hide code; set to TRUE to see code
  fig.pos = 'th', # Places figures at top or here
  out.width = '100%', dpi = 300, # Figure resolution and size
  fig.env="figure"
) # Latex figure environment

options(knitr.table.format = "latex") # For kable tables to write LaTeX table directly
```

@@

@@

```{r, echo = FALSE, out.height = '75%', fig.align = 'center'}
knitr::include_graphics("flowchart.png")
```

@@

```{r kabletable, echo=FALSE}
df = data.frame(date = c("2000-01-01",
                         "...",
                         "2000-07-01",
                         "...",
                         "2000-12-31"),
                temperature= c(2.51,
                        "...",
                        16.32,
                        "...",
                        3.13
                        ))

# kable can alse be used for creating tables
knitr::kable(df,
             booktabs=TRUE)
```

@@

```{r, echo = TRUE}
# create date object for a year with 30 min reading intervals
dates <- seq(from = ymd_hms("2000-07-01 00:00:00"),
             to = ymd_hms("2001-06-30 23:59:59"), length.out = 17568)

@@

```{r, echo = TRUE}
plot_check_temp(data = year_sim,
                dates = date,
                temperature = temp,
                temp_min = 0, # temp_min and max lines are 
                              # user customizable
                temp_max = 25)
```

@@

```{r, echo = TRUE}
# summarize
year_sim_summ <- summarize_temp(data = year_sim,
                                dates = date,
                                temperature = temp)

@@

Users can either use Salmonid models from `model_table` included in the package and the Shiny app. As discussed, these models are included because parameterizations already existed in the literature. However, these parameterizations are limited to Pacific Salmon and Bull Trout. To widen the application of the effective value approach, we include a `fit_model()` function, which is species agnostic (as long as development generally follows the power law).

The function `fit_model()` uses data in which average incubation temperature (°C) and days to phenological event (as two vectors) are the inputs. The function estimates parameter coefficients for *log~e~a* and *b* using `stats::nls()`

@@

Below, we demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package. We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for fishes beyond the scope of the original effective value approach. These parameterizations are for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*) from @webster1948, Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005. We provide parameterization for just Smallmouth Bass here for concision, but the full code set for all species is available at <https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper.Rmd>.

@@

```{r, echo = TRUE}
###  make temp regime
set.seed(123)
# create random temps and corresponding dates
temps_sim <- sort(rnorm(n =30, mean = 16, sd = 1), decreasing = FALSE)
dates_sim <-  seq(from = ymd("2000-07-01"),
             to = ymd("2000-07-31"), length.out = 30)

@@

```{r, echo = TRUE}
#model fits
smb_mod$r_squared; cat_mod$r_squared; sturgeon_mod$r_squared

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE}
### predict_phenology

@@

```{r, echo = TRUE}
sockeye_hatch_mod <- model_select(
  author = "Beacham and Murray 1990", 
  species = "sockeye", 
  model = 2, 
  development_type = "hatch"
  )
```

@@

```{r, warning=FALSE, echo = TRUE}
WI_hatch <- predict_phenology(
  data = woody_island,
  dates = date,
  temperature = temp_c,
  spawn.date = "1990-08-18",
  model = sockeye_hatch_mod
  )
```

@@

```{r, echo = TRUE}
WI_hatch$days2done
WI_hatch$dev.period
```

@@

```{r, eval=FALSE, echo = TRUE}
str(WI_hatch)
```

@@

`{r"} plot_phenology(WI_hatch)`

@@

A common management scenario where developmental phenology might be useful would be trying to understand if fish might be free-moving before some management action. For instance, will fish have emerged from redds when a stream section has been opened to grazing or road work?

In this scenario, we will consider the road work example and Bull Trout, a threatened fish in the United States under the Endangered Species Act [@nolfi2024], and the Crooked River, a key Bull Trout population in the Boise River watershed. In this hypothetical scenario, the Forest Fisheries Biologist wants to know if fish will likely be out of the gravel and free-swimming by June 1st as Bull Trout are particularly sensitive to sediment. In this system, it is expected that Bull Trout will be done spawning by the end of September, so we'll consider the last possible spawn date as September 30th.

We demonstrate this first case study using the graphical user interface portion of the **hatchR** ecosystem found at <https://elifelts.shinyapps.io/hatchR/>. Users will first upload their data with the `Import Data` window, which requires them to select their file on their personal computer, provide the program with the columns corresponding for dates and temperatures, and then provide the format in which dates are coded (*e.g.*, year-month-day or day-month-year). Data used in this case study are from the `crooked_river` data set from R and can either be written out from the package or accessed at <https://github.com/bmait101/hatchR/blob/master/extdata/crooked_river.csv/>. Once data is uploaded the program automatically plots the user's data using `plot_check_temp()` in the background and provides them the outputted graphical check. After uploading and checking data, the user switches to the `Model Phenology` window. In this circumstance, we use the preloaded parameterization for bull trout from @austin2019 with the `Existing` button for model selection, which the user selects with the various drop down options in the menu. After the model is selected, the user can choose multiple spawn dates from the interactive calendar provided. We show results for spawning for September 30th as indicated in the example above. Once dates are chosen, a table entry for each spawn date is outputted in the `Phenology Summaries` tab and corresponding plot with data from each spawn date in the `Timeline Plot` tab. Output from predicting phenology and the resulting figure are downloadable from their respective tabs. The process is demonstrated in full in Figure XXX, but the interface is described more completely on **hatchR**'s Shiny website.

@@

```{r, echo = TRUE}
hatch_res <- isaak_summ_bt |> 
  mutate(
    dev_period = map2(
      summ_obj, spawn_dates, # map across our site object and spawn dates
      predict_phenology,
      temperature = daily_temp,
      model = bt_hatch,
      dates = date
      ) |> 
      map_df("dev.period") |> # pull out just dev.period results
      list()
    ) |> 
  select(site, dev_period) |> # just select the columns we want
  unnest(cols = c(dev_period)) |> # unnest everything
  mutate(days_to_hatch = stop - start) # make a new column of days to hatch
```

```{r, message=FALSE, echo=FALSE}
cut_ints <- date(
  c(
    "2011-08-31", "2012-08-31",
    "2013-08-31", "2014-08-31",
    "2015-08-31"
  )
)

@@
